---
title: "Non-B DNA Motif Analysis with `nonBgfa`"
author: "Saifun Nahar"
format:
  html:
    code-fold: true
    toc: true
---

## Introduction

This notebook demonstrates the usage of the refactored `nonBgfa` package to find non-B DNA forming motifs in genomic sequences. The underlying engine is an optimized C library refactored from the original GFA suite.

## Setup

First, we install (if necessary) and load the `nonBgfa` package.

```{r}
#| label: setup
#| message: false
# Ensure the package is installed from the local source
# devtools::install(".") 
library(nonBgfa)
library(ggplot2)
```

## Analysis in R

We will read the test sequence and find motifs using the default parameters.

```{r}
#| label: analysis-r
# Helper to read fasta
read_fasta_string <- function(file) {
  lines <- readLines(file)
  seq <- paste(lines[!grepl("^>", lines)], collapse = "")
  return(seq)
}

# Load test data
# (Assuming gfa_test.fasta is extracted from test_files.tar)
fasta_file <- "gfa_test.fasta"
if (!file.exists(fasta_file)) {
  system("tar -xf test_files.tar gfa_test.fasta")
}

sequence <- read_fasta_string(fasta_file)
results <- findMotifs(sequence)

# Summary of found motifs
summary_df <- data.frame(
  MotifType = names(results),
  Count = sapply(results, nrow)
)

print(summary_df)
```

### Visualizing Motif Distribution

```{r}
#| label: plot-r
#| fig-cap: "Distribution of found non-B motifs along the sequence"

# Combine all results for plotting
all_motifs <- do.call(rbind, lapply(names(results), function(n) {
  df <- results[[n]]
  if(nrow(df) > 0) {
    df$Type <- n
    return(df[, c("start", "end", "Type")])
  }
  return(NULL)
}))

ggplot(all_motifs, aes(x = start, xend = end, y = Type, yend = Type, color = Type)) +
  geom_segment(size = 2) +
  theme_minimal() +
  labs(title = "Location of Non-B DNA Motifs", x = "Position (bp)", y = "")
```

## Analysis in Python

Python users can interface with the R package using the `reticulate` library.

```{python}
#| label: analysis-python
import pandas as pd

# This assumes 'reticulate' is configured to allow Python to call R
# and that 'nonBgfa' is available in the R environment.
try:
    from rpy2.robjects.packages import importr
    from rpy2.robjects import r, globalenv
    
    # Import the R package
    nonBgfa = importr('nonBgfa')
    
    # Call the R function from Python
    # We can pass the sequence from R to Python or vice versa
    seq_python = r('sequence')
    py_results = r('findMotifs(sequence)')
    
    print("Python received results for:", py_results.names)
    # Convert one to a pandas DataFrame if needed
    # ir_df = pd.DataFrame(py_results.rx2('IR'))
    # print(ir_df.head())
except ImportError:
    print("rpy2 not installed, skipping Python integration demo.")
```

## Conclusion

The `nonBgfa` package provides a high-performance, cross-language interface for genomic feature analysis. The refactored memory strategy ensures it can handle large sequences efficiently within modern data science workflows.
